Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-08-20 latex/main.tex
Updates from Overleaf

ancestral recombination graphs, recombination, tree sequence, SARS-CoV-2

@@

Sc2ts uses collection dates to build a time-resolved ARG. 
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
likely due to errors in their collection dates 
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
which is estimated to have emerged in November, 2021; REF). 
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”. 
We define the “HMM cost” of a sample to be the total penalty cost for switches and mismatches (mutations) in its copying path. 
We parameterize the LS model such that each mismatch has a penalty cost of one 
and a switch has a penalty cost of \textit{k} (here, set to 4), 
which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}). 
If the HMM cost of a sample exceeds a threshold (here, set to 7), sc2ts does not immediately add the sample to the ARG.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{methodology.png}
    \caption{Schematic overview of sc2ts. (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. Sample 0 is used to initialize the ARG as the root. Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child. The resulting ARG contains two local trees, one on each side of the inferred breakpoint. (B) Copying paths for Sample 4 inferred under the LS model while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).}
    \label{fig:methodology}
\end{figure}

@@

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
on the basis of the HMM and available data. 
For all these Pango Xs, the number of recombination-informative sites 
fell below our threshold of detection (i.e. k = 4 in the HMM). 
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
For example, the closely related XN and XAU can be explained 
by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}). 
%%% TODO: Say something about XAJ and XAN.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
because the HMM treated characteristic deletions as missing data.

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020} 
to relate the timing of recombination events with global case numbers. 
Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 and 
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{recombinants_over_time}). 
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
Alongside the case numbers, the black curve shows the number of recombinants inferred within each week along the ARG, 
focusing on the 386 “robust” events. 
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high. 
Overall, the number of recombinant events rose with the global number of cases 
(Figure 5B; linear regression: 1.2 + 2.9 10^{-7} x, p = 3.2  10^{-7}). 
Recombination events also rose with the diversity of lineages in each week 
(Figure 5C; linear regression: 1.0 + 5.3 x, p = 1.3 10^{-7}), 
measured as the chance that two randomly drawn viruses have different Scorpio designations 
(H, referred to as the “expected heterozygosity”).

We then asked which single measure best explains the number of recombination events inferred each week, 
considering \textit{n}, \textit{H}, \textit{n H}, \textit{n}^2, \textit{n}^2 \textit{H}. 
We included the squared number of cases each week thinking that it might better predict co-infections. 
The best single predictor of recombination events involved the product of cases and expected heterozygosity (\textit{n H}), 
with an adjusted R^2 of 25.3\%. 
This measure of case diversity (\textit{n H}) explained 62\% more of the variation in recombination events per week 
than the next best predictor (\textit{H}, adjusted R^2 of 15.7\%). 
While consistent with the hypotheses that recombination events occur more often when infection rates are common 
and are detected more often when viral diversity is high, we note several sources of uncertainty. 
Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}. 
Furthermore, there may be substantial uncertainty in the timing of a recombination event, 
particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022}) 
with substantial delays before onward cases are detected. 
Finally, the filters that we applied removed all but the most strongly supported recombination events, 
although similar conclusions were obtained when all 929 recombinants were considered (STAR Methods).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
    \caption{Recombination events across the pandemic. (A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio labels, with the key showing those groupings with \geq 1,000 samples in the Viridian dataset. Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity (C); solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
    \label{fig:recombinants_over_time}
\end{figure}

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale, 
which overwhelmed existing computational approaches, 
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}. 
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct, 
the functionality required largely overlaps. 
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024} 
and the VCF Zarr data storage format \cite{Czech2025} 
to describe relationships among virus samples, building an ARG over time as data becomes available. 
By reusing high-quality and high-performance components of these toolkits, 
we could develop the sc2ts inference method itself and 
perform all the pandemic-scale analyses in this article with Python code alone and 
no compiled languages.

To illustrate the power of these libraries we provide some illustrative examples 
in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments. 
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
represents 31.34 GiB of nucleotide calls. 
Of the 80,574,518 missing data calls (Ns) in the alignments, 
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%). 
Additionally, 413,303 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code). 
UShER imputed only 4 calls from this set incorrectly, 
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations. 
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 Gb 
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
Performing this analysis using matUtils would require exporting the full dataset to VCF, 
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments. 
In contrast with matUtils and most phylogenetic software, 
which provide a command line interface and suite of predefined operations that typically output text files, 
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}. 
As an illustration of the open-ended and efficient nature of the analyses this enables, 
we computed the number of polytomies (nodes that have > 2 children) 
and the maximum number of children per node in the Usher and sc2ts outputs. 
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148) 
and compute the maximum number of children (5918). 
In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and 
maximum number of children (7708) across all 348 local trees.

@@

We present sc2ts, a real-time ARG inference method, 
and applied it to build a ~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
By identifying specific samples as parents, 
sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods, 
which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages. 
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022} 
which detects recombinants post hoc by searching for unusual branches in a given phylogeny, 
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution. 
Combined with new methods and tools to evaluate evidence for recombination, 
sc2ts is a significant technological advance over current methods for the discovery 
and analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2. 
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution, 
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}. 
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER. 
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes, 
allowing for convenient and efficient reuse of this valuable dataset 
for a variety of evolutionary and genetic analyses beyond those presented here. 
We hope that the ARG will spur on new research areas, in which recombination is naturally and fully accounted for, 
and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022} 
has made on our understanding of human genetic ancestry and demographic past.

We contribute several novel methods and concepts which may find applications beyond sc2ts. 
First, we devise a simplified LS model, which extends the classic concept of parsimony. 
In sc2ts, this model allows for accelerated HMM likelihood calculations, 
enabling reconstruction of the pandemic-scale ARG presented here. 
Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup. 
Second, we propose the concept of mutations averted by recombination, 
which leads to an interpretable quantity to weigh 
the support for the hypothesis that a given sample is a recombinant 
against the hypothesis that it is a non-recombinant, 
considering the sampled genetic diversity in an ARG and a specific HMM. 
Third, we propose the concept of HMM cost to identify poorly explained samples 
which are either too divergent or have a high number of mutations inconsistent with their collection dates. 
Here we use HMM cost to reject time travelling samples, 
which no existing phylogenetic method detects and handles during inference 
(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}), 
resulting in a clean dataset freed of samples with erroneous collection dates that may affect downstream analyses. 
Future work may be estimating dates for the suspected time travelling samples using the ARG provided here 
and the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
but it has been understudied partly owing to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
While the methods and tools developed here are tailored for SARS-CoV-2, 
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is 
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops. 
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem, 
making them highly adaptable for a broad range of applications. 
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers, 
enabling and accelerating studies of SARS-CoV-2 for years to come.

@@

\subsubsection*{Description of sc2ts}
Sc2ts (pronounced “scoots”, optionally) 
is a method for inferring Ancestral Recombination Graphs (ARGs) 
from densely sampled genome sequence data in real time, 
in which recombination occurs at a low but significant rate. 
The basic idea is to incrementally update an ARG regularly 
with the sequences collected since the previous batch (Figure 1A). 
We used daily batches for SARS-CoV-2, 
and use days as the time interval for clarity in what follows. 
The first step is to find likely “copying paths” under the Li and Stephens (LS) model \cite{Li2003} 
for each sequence in the daily batch to the current ARG. 
These copying paths will mostly consist of a new sample sequence copying 
from a node in the ARG with a small number of mutations, 
and often many samples from a daily batch will copy from the same ARG node. 
The second step is then to “resolve” these implied polytomies 
by using a standard neighbor-joining algorithm \cite{Saitou1987}. 
This greedy update strategy inevitably leads to unparsimonious topologies, 
and the third step is to then increase the overall parsimony of the inferred ARG 
by making some simple topological updates. 
Recombination is inferred as an integral and ongoing part of this process, 
requiring only a few additional steps to facilitate later analysis. 
The result of inference for a given day is then a genealogy recording genetic inheritance 
as well as mutation and recombination events 
for all of the samples added to the ARG up to that day.

\subsubsection*{Ancestral Recombination Graphs}
%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997) 
to define an alternative formulation of the coalescent with recombination (Hudson, 1983), 
the term “Ancestral Recombination Graph” (ARG) has come to be used 
more generally to describe not just realisations of the model, 
but any recombinant genetic ancestry (Minichiello and Durbin, 2006). 
While there is some subtlety in the details \cite{Wong2024}, 
we can think of an ARG as being any graph 
that encodes the reticulate genetic genealogy of a sample of sequences 
under the influence of recombination.

The “succinct tree sequence” (STS) is an ARG data structure 
that is both general (in terms of the types of ancestry that can be described) 
and computationally efficient \cite{Wong2024}. 
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016}, 
STS has been extended and applied to forward-time simulations \cite{Kelleher2018,Haller2019}, 
calculation of population genetics statistics \cite{Ralph2020} 
and ARG inference \cite{Kelleher2019,Wohns2022}. 
The STS is based on a simple tabular representation, 
which defines a set of nodes, edges, sites and mutations. 
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor. 
The genetic inheritance between a pair of nodes along a segment of genome 
is defined by the edge (ℓ, r, p, c), 
which states that child node c inherited its genome from parent node p 
from left coordinate ℓ to right coordinate r. 
A site defines a position on the genome and the ancestral state (allele) at that site. 
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

Given the node, edge, site and mutation tables, 
we can efficiently construct the local genealogical trees along the genome (arising from recombination) 
and perform a range of calculations efficiently 
by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}. 
These algorithms have led to performance increases of several orders of magnitude over 
the state-of-the-art in a range of applications \cite{Kelleher2016,Kelleher2018,Kelleher2019,Ralph2020,Baumdicker2022}. 
The STS encoding is also very concise, allowing, for example, 
for millions of complete human genomes to be stored in a few Gigabytes of space \cite{Kelleher2019}, 
and so it is particularly well suited to representing large SARS-CoV-2 genealogies.

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analysed 
using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}. 
Tskit is widely used in population genetics, and has been incorporated into many downstream applications 
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).

\subsubsection{Li and Stephens haplotype-copying model}
LS model \cite{Li2003} is an approximation of the coalescent with recombination 
that captures many of the key features of the joint processes of mutation and recombination. 
It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}. 
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states), 
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states). 
The process of switching between the genomes in the reference panel is governed by a transition matrix, 
and mismatches between a focal genome and a genome of the reference panel (from which it is “copied”) are permitted as emissions.

In the original formulation of the LS model (see Appendix S1 in Li and Stephens, 2003), 
the probability of switching between the hidden states (i.e., genomes in a reference panel) 
when going from one site to the next is dependent on 
(1) the rate of recombination between the sites and 
(2) the number of genomes in the reference panel (\textit{n}). 
Sc2ts uses the same formulation except that n is set to 1. 
This special case of the LS model leads to an intuitive interpretation of the ratio 
between the probability of switching and the probability of mismatch: 
When considering only two copying paths, the path having k mismatches 
and the other path having one switch but no mismatch are equally probable. 
We call this parameter k the “recombination penalty”. 
This parameter thus allows us to adjust the relative importance of recombination and mutation 
when finding the best explanation for a given sample. 
For simplicity, we assume that any single base mutation causing a mismatch is equally probable, 
ignoring transition/transversion biases and variation of mutation rate along the genome. 
As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.

>>>>

ancestral recombination graphs, population genetics, recombination, tree sequence, SARS-CoV-2

@@

Sc2ts uses collection dates to build a time-resolved ARG. 
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
likely due to errors in their collection dates 
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
which is estimated to have emerged in November, 2021; REF). 
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”. 
We define the “HMM cost” of a sample to be 
the total penalty cost for switches and mismatches (mutations) in its copying path. 
We parameterize the LS model such that each mismatch has a penalty cost of one 
and a switch has a penalty cost of $k$ (here, set to 4), 
which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}). 
If the HMM cost of a sample exceeds a threshold (here, set to 7), 
sc2ts does not immediately add the sample to the ARG.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{methodology.png}
    \caption{Schematic overview of sc2ts. (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. Sample 0 is used to initialize the ARG as the root. Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child. The resulting ARG contains two local trees, one on each side of the inferred breakpoint. (B) Copying paths for Sample 4 inferred under the LS model while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).}
    \label{fig:methodology}
\end{figure}

@@

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
on the basis of the HMM and available data. 
For all these Pango Xs, the number of recombination-informative sites 
fell below our threshold of detection (i.e. $k = 4$ in the HMM). 
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
For example, the closely related XN and XAU can be explained 
by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}). 
%%% TODO: Say something about XAJ and XAN.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
because the HMM treated characteristic deletions as missing data.

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020} 
to relate the timing of recombination events with global case numbers. 
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and 
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{recombinants_over_time}). 
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
Alongside the case numbers, 
the black curve shows the number of recombinants inferred within each week along the ARG, 
focusing on the 386 “robust” events. 
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high. 
Overall, the number of recombinant events rose with the global number of cases 
(Figure 5B; linear regression: $1.2 + 2.9 10^{-7} x$, $p = 3.2e^{-7}$). 
Recombination events also rose with the diversity of lineages in each week 
(Figure 5C; linear regression: $1.0 + 5.3 x$, $p = 1.3e^{-7}$), 
measured as the chance that two randomly drawn viruses have different Scorpio designations 
($H$, referred to as the “expected heterozygosity”).

We then asked which single measure best explains the number of recombination events inferred each week, 
considering $n$, $H$, $n H$, $n^2$, $n^2 H$. 
We included the squared number of cases each week thinking that it might better predict co-infections. 
The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n H$), 
with an adjusted $R^2$ of 25.3\%. 
This measure of case diversity ($n H$) explained 62\% more of the variation in recombination events per week 
than the next best predictor ($H$, adjusted $R^2$ of 15.7\%). 
While consistent with the hypotheses that recombination events occur more often when infection rates are common 
and are detected more often when viral diversity is high, we note several sources of uncertainty. 
Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}. 
Furthermore, there may be substantial uncertainty in the timing of a recombination event, 
particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022}) 
with substantial delays before onward cases are detected. 
Finally, the filters that we applied removed all but the most strongly supported recombination events, 
although similar conclusions were obtained when all 929 recombinants were considered (STAR Methods).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
    \caption{Recombination events across the pandemic. (A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio labels, with the key showing those groupings with $\geq 1,000$ samples in the Viridian dataset. Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity (C); solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
    \label{fig:recombinants_over_time}
\end{figure}

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale, 
which overwhelmed existing computational approaches, 
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}. 
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct, 
the functionality required largely overlaps. 
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024} 
and the VCF Zarr data storage format \cite{Czech2025} 
to describe relationships among virus samples, building an ARG over time as data becomes available. 
By reusing high-quality and high-performance components of these toolkits, 
we could develop the sc2ts inference method itself and 
perform all the pandemic-scale analyses in this article with Python code alone and 
no compiled languages.

To illustrate the power of these libraries, we provide some examples 
in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments. 
The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
represents 31.34 GB of nucleotide calls. 
Of the 80,574,518 missing data calls (Ns) in the alignments, 
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%). 
Additionally, 413,303 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code). 
UShER imputed only 4 calls from this set incorrectly, 
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations. 
This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB 
(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
Performing this analysis using matUtils would require exporting the full dataset to VCF, 
and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments. 
In contrast with matUtils and most phylogenetic software, 
which provide a command line interface and suite of predefined operations that typically output text files, 
tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}. 
As an illustration of the open-ended and efficient nature of the analyses this enables, 
we computed the number of polytomies (nodes that have > 2 children) 
and the maximum number of children per node in the Usher and sc2ts outputs. 
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148) 
and compute the maximum number of children (5918). 
In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and 
maximum number of children (7708) across all 348 local trees.

@@

We present sc2ts, a real-time ARG inference method, 
and applied it to build a \~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
By identifying specific samples as parents, 
sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods, 
which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages. 
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022} 
which detects recombinants post hoc by searching for unusual branches in a given phylogeny, 
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution. 
Combined with new methods and tools to evaluate evidence for recombination, 
sc2ts is a significant technological advance over current methods for the discovery and 
analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2. 
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution, 
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}. 
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER. 
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes, 
allowing for convenient and efficient reuse of this valuable dataset 
for a variety of evolutionary and genetic analyses beyond those presented here. 
We hope that the ARG will spur on new research areas, 
in which recombination is naturally and fully accounted for, 
and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022} 
has made on our understanding of human genetic ancestry and demographic past (e.g., REF).

We contribute several novel methods and concepts which may find applications beyond sc2ts. 
First, we devise a simplified LS model, which extends the classic concept of parsimony. 
In sc2ts, this model allows for accelerated HMM likelihood calculations, 
enabling reconstruction of the pandemic-scale ARG presented here. 
Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup. 
Second, we propose the concept of mutations averted by recombination, 
which leads to an interpretable quantity to weigh 
the support for the hypothesis that a given sample is a recombinant 
against the hypothesis that it is a non-recombinant, 
considering the sampled genetic diversity in an ARG and a specific HMM. 
Third, we propose the concept of HMM cost to identify poorly explained samples 
which are either too divergent or have a high number of mutations inconsistent with their collection dates. 
Here we use HMM cost to reject time traveling samples, 
which no existing phylogenetic method detects and handles during inference 
(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}), 
resulting in a clean dataset rid of samples with erroneous collection dates that may affect downstream analyses. 
Future work may be estimating dates for the suspected time traveling samples using the ARG provided here 
and the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
While the methods and tools developed here are tailored for SARS-CoV-2, 
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is 
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops. 
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem, 
making them highly adaptable for a broad range of applications. 
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers, 
enabling and accelerating studies of SARS-CoV-2 for years to come.

@@

\subsubsection*{Description of sc2ts}

Sc2ts (pronounced “scoots”, optionally) 
is a method for inferring Ancestral Recombination Graphs (ARGs) 
from densely sampled genome sequence data in real time, 
in which recombination occurs at a low but significant rate. 
The basic idea is to incrementally update an ARG regularly 
with the sequences collected since the previous batch \ref{fig:methodology}. 
We used daily batches for SARS-CoV-2, 
and use days as the time interval for clarity in what follows. 
The first step is to find likely “copying paths” under the Li and Stephens (LS) model \cite{Li2003} 
for each sequence in the daily batch to the current ARG. 
These copying paths will mostly consist of a new sample sequence copying 
from a node in the ARG with a small number of mutations, 
and often many samples from a daily batch will copy from the same ARG node. 
The second step is then to “resolve” these implied polytomies 
by using a standard neighbor-joining algorithm \cite{Saitou1987}. 
This greedy update strategy inevitably leads to unparsimonious topologies, 
and the third step is to then increase the overall parsimony of the inferred ARG 
by making some simple topological updates. 
Recombination is inferred as an integral and ongoing part of this process, 
requiring only a few additional steps to facilitate later analysis. 
The result of inference for a given day is then a genealogy recording genetic inheritance 
as well as mutation and recombination events 
for all of the samples added to the ARG up to that day.

\subsubsection*{Ancestral Recombination Graphs}

%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997) 
to define an alternative formulation of the coalescent with recombination (Hudson, 1983), 
the term “Ancestral Recombination Graph” has come to be used 
more generally to describe not just realizations of the model, 
but any recombinant genetic ancestry (Minichiello and Durbin, 2006). 
While there is some subtlety in the details \cite{Wong2024}, 
we can think of an ARG as being any graph 
that encodes the reticulate genetic genealogy of a sample of sequences 
under the influence of recombination.

The “succinct tree sequence” (STS) is an ARG data structure 
that is both general (in terms of the types of ancestry that can be described) 
and computationally efficient \cite{Wong2024}. 
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016}, 
STS has been extended and applied to 
forward-time simulations \cite{Kelleher2018,Haller2019}, 
calculation of population genetics statistics \cite{Ralph2020} and 
ARG inference \cite{Kelleher2019,Wohns2022}. 
The STS is based on a simple tabular representation, 
which defines a set of nodes, edges, sites and mutations. 
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor. 
The genetic inheritance between a pair of nodes along a segment of genome 
is defined by the edge ($l$, $r$, $p$, $c$), 
which states that child node $c$ inherited its genome from parent node $p$ 
from left coordinate $l$ to right coordinate $r$. 
A site defines a position on the genome and the ancestral state (allele) at that site. 
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

Given the node, edge, site and mutation tables, 
we can efficiently construct the local genealogical trees along the genome (arising from recombination) 
and perform a range of calculations efficiently 
by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}. 
These algorithms have led to performance increases of several orders of magnitude over 
the state-of-the-art in a range of applications \cite{Kelleher2016,Kelleher2018,Kelleher2019,Ralph2020,Baumdicker2022}. 
The STS encoding is also very concise, allowing, for example, 
for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019}, 
and so it is particularly well suited to representing large SARS-CoV-2 genealogies.

The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed 
using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}. 
Tskit is widely used in population genetics, and has been incorporated into many downstream applications 
(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).

\subsubsection{Li and Stephens haplotype-copying model}

LS model \cite{Li2003} is an approximation of the coalescent with recombination 
that captures many of the key features of the joint processes of mutation and recombination. 
It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}. 
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states), 
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states). 
The process of switching between the genomes in the reference panel is governed by a transition matrix, 
and mismatches between a focal genome and a genome of the reference panel (from which it is “copied”) are permitted as emissions.

In the original formulation of the LS model (see Appendix S1 of \cite{Li2003}), 
the probability of switching between the hidden states (i.e., genomes in a reference panel) 
when going from one site to the next is dependent on 
(1) the rate of recombination between the sites and 
(2) the number of genomes in the reference panel ($n$). 
Sc2ts uses the same formulation except that n is set to 1. 
This special case of the LS model leads to an intuitive interpretation of the ratio 
between the probability of switching and the probability of mismatch: 
When considering only two copying paths, the path having k mismatches 
and the other path having one switch but no mismatch are equally probable. 
We call this parameter k the “recombination penalty”. 
This parameter thus allows us to adjust the relative importance of recombination and mutation 
when finding the best explanation for a given sample. 
For simplicity, we assume that any single base mutation causing a mismatch is equally probable, 
ignoring transition-transversion biases and variation of mutation rate along the genome. 
As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.

In our early attempts to build ARGs, we explored using several values of $k$. 
We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}}(REF). 
We would expect that a good value for $k$ should lead to parsimonious ARGs 
that capture plausible recombination events without incurring too many mutations, particularly reversions. 
A value of $k$ too low could result in many artifactual recombination events, 
whereas a value of $k$ too high could miss many genuine recombination events. 
We found that ARGs built with $k$ set to 3 contained many dubious recombination events, 
but ARGs built with $k$ set to 5 failed 
to include the well supported recombination events documented in a previous study \cite{Jackson2021}. 
Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.
